New cardinality estimation algorithms for HyperLogLog sketches

### 2016-10-26 paper/paper.tex
switched from gsl to dlib BFGS algorithm

For the general case the Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm \cite{Press2007} can be applied, which is a very popular algorithm for non-linear optimization problems. In particular, we used the implementation provided by the GNU scientific library \cite{Galassi2016}.

@@

We also investigated, if the new approach could give better estimates for the union operation. The conventional approach merges both HyperLogLog sketches using \eqref{equ:hll_union} and estimates the union size using single sketch cardinality estimation. The joint cardinality estimation algorithm opens another way. The union can also be estimated by simply summing up the three estimates for the sets $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$. The corresponding results are shown in \cref{tbl:joint_estimation_abx}. As can be clearly seen the joint estimation algorithm is also able to improve the cardinality estimation of unions by an significant amount.

\begin{table}
\centering
\caption{List of true cardinalities of the pair-wise disjoint sets $\symSetA$, $\symSetB$, and $\symSetX$, for which maximum likelihood estimation from two HyperLogLog sketches with parameters $\symPrecision=20$ and $\symRegRange=44$ representing sets $\symSetA\cup\symSetX$ and  $\symSetB\cup\symSetX$ was investigated. The average number of function \eqref{equ:max_likelihood_phi_func} and gradient evaluations \eqref{equ:df_dphi} are also given as needed by our implementation of \cref{alg:joint_cardinality_estimation} that uses the BFGS algorithm.
}
\label{tbl:joint_estimation_cases}
\csvreader[before reading=\scriptsize, tabular=r|rrr|rr|rr, table head= & 
\multicolumn{3}{c|}{true cardinalities} &
\multicolumn{1}{c}{Jaccard index} &
\multicolumn{1}{c|}{cardinality ratio} &
\multicolumn{2}{c}{avg num evaluations}
\\
\# &  
\multicolumn{1}{c}{$|\symSetA|$} & 
\multicolumn{1}{c}{$|\symSetB|$} & 
\multicolumn{1}{c|}{$|\symSetX|$} & 
\multicolumn{1}{c}{$\frac{|\symSetX|}{|\symSetA\cup\symSetB\cup\symSetX|}$} & 
\multicolumn{1}{c|}{$\frac{|\symSetA\cup\symSetX|}{|\symSetB\cup\symSetX|}$} &
\multicolumn{1}{c}{$\symFunc$} &
\multicolumn{1}{c}{$\nabla\symFunc$}
\\\hline]
{joint_cardinality_calculation.csv}{
trueCardA=\trueCardA,
trueCardB=\trueCardB,
trueCardX=\trueCardX,
trueJaccardIdx=\trueJaccardIdx,
trueRatio=\trueRatio,
avgNumFunctionEvaluations=\avgNumFunctionEvaluations,
avgNumGradientEvaluations=\avgNumGradientEvaluations,
}{
\thecsvrow & 
\num{\trueCardA} & 
\num{\trueCardB} & 
\num{\trueCardX} & 
\numformat{\trueJaccardIdx} & 
\numformattwo{\trueRatio} &
\numformattwo{\avgNumFunctionEvaluations} &
\numformattwo{\avgNumGradientEvaluations}
}
\end{table}

>>>>

For the general case the Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm \cite{Press2007} can be applied, which is a very popular algorithm for non-linear optimization problems. In particular, we used the implementation provided by the Dlib C++ library \cite{King2009}.

@@

We also investigated, if the new approach could give better estimates for the union operation. The conventional approach merges both HyperLogLog sketches using \eqref{equ:hll_union} and estimates the union size using single sketch cardinality estimation. The joint cardinality estimation algorithm opens another way. The union can also be estimated by simply summing up the three estimates for the sets $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$. The corresponding results are shown in \cref{tbl:joint_estimation_abx}. As can be clearly seen the joint estimation algorithm is also able to improve the cardinality estimation of unions by a significant amount.

\begin{table}
\centering
\caption{List of true cardinalities of the pair-wise disjoint sets $\symSetA$, $\symSetB$, and $\symSetX$, for which maximum likelihood estimation from two HyperLogLog sketches with parameters $\symPrecision=20$ and $\symRegRange=44$ representing sets $\symSetS_1 = \symSetA\cup\symSetX$ and  $\symSetS_2 = \symSetB\cup\symSetX$ was investigated. The last column shows the average number of needed log-likelihood function 
 \eqref{equ:max_likelihood_phi_func} and gradient evaluations \eqref{equ:df_dphi} until convergence of the BFGS algorithm.
}
\label{tbl:joint_estimation_cases}
\csvreader[before reading=\scriptsize, tabular=r|rrr|rr|rr, table head= & 
\multicolumn{3}{c|}{true cardinalities} &
\multicolumn{1}{c}{Jaccard index} &
\multicolumn{1}{c|}{cardinality ratio} &
\multicolumn{1}{c}{avg number}
\\
\# &  
\multicolumn{1}{c}{$|\symSetA|$} & 
\multicolumn{1}{c}{$|\symSetB|$} & 
\multicolumn{1}{c|}{$|\symSetX|$} & 
\multicolumn{1}{c}{$|\symSetS_1\cap\symSetS_2|/|\symSetS_1\cup\symSetS_2|$} & 
\multicolumn{1}{c|}{$|\symSetS_1|/|\symSetS_2|$} &
\multicolumn{1}{c}{iterations}
\\\hline]
{joint_cardinality_calculation.csv}{
trueCardA=\trueCardA,
trueCardB=\trueCardB,
trueCardX=\trueCardX,
trueJaccardIdx=\trueJaccardIdx,
trueRatio=\trueRatio,
avgNumEvaluations=\avgNumEvaluations
}{
\thecsvrow & 
\num{\trueCardA} & 
\num{\trueCardB} & 
\num{\trueCardX} & 
\numformat{\trueJaccardIdx} & 
\numformattwo{\trueRatio} &
\numformattwo{\avgNumEvaluations}
}
\end{table}
